---
title: "Seven Steps to Reproducible (Computational) Research"
author: "Liz Ing-Simmons"
format: revealjs
---

## What is reproducibility?

* definitions [TODO]
* add images 
* we'll mainly consider ...

## Why does reproducibility matter?

* reproducibility crisis
* research should be trustworthy
* avoid waste - wasted time regenerating data, or waste time building on work that isn't reproducible

## Seven steps

1. Planning to be organised
2. Keeping data tidy
3. Documentation
4. Methodology and protocols
5. Testing and controls
6. Automation
7. Publishing

## Planning to be organised

Before you start a project:
* What data will you collect?
* What code will you write?

## Planning to be organised

How will you organise your data / code?
* Directory structure
* File naming conventions
  * e.g. project-specific abbreviations
  * file names that are both human-readable and computer-readable

## Planning to be organised: tools 🔧

"Data Management Plans" are required by many[?] funders.
Structure to document what data you will collect, where it will be stored, who is responsible
Good opportunity to think about and document these things - and think about how data will be handled at end of the project (what can be deleted, what can be shared)

## Planning to be organised: tools 🔧

Project templates [TODO examples]

## Keeping data tidy

::: notes
(this refers to text-based data inside files)
:::

*"each variable is a column, each observation is a row, and each type of observational unit is a table"*

[Wickham, 2014](http://www.jstatsoft.org/v59/i10/)

. . .

* Tidy data facilitates analysis
* Makes data cleaning easier
* Works well with statistical tools

## Keeping data tidy

*"tidy datasets are all alike but every messy dataset is messy in its own way"*

. . .

:::: {.columns}

::: {.column width="50%"}
```{r}
library(tibble)
classroom <- tribble(
  ~name,    ~quiz1, ~quiz2, ~exam1,
  "Billy",  NA,     "D",    "C",
  "Suzy",   "F",    NA,     NA,
  "Lionel", "B",    "C",    "B",
  "Jenny",  "A",    "A",    "B"
  )
classroom
```
:::

::: {.column width="50%"}

```{r}
tribble(
  ~assessment, ~Billy, ~Suzy, ~Lionel, ~Jenny,
  "quiz1",     NA,     "F",   "B",     "A",
  "quiz2",     "D",    NA,    "C",     "A",
  "exam1",     "C",    NA,    "B",     "B"
  )
```
:::

::::

. . .

* Compact and convenient for humans, but not for computers


## Keeping data tidy

:::: {.columns}

::: {.column width="40%"}
```{r}
library(tidyr)
library(dplyr)

classroom %>% 
  pivot_longer(quiz1:exam1, names_to = "assessment", values_to = "grade") %>% 
  arrange(name, assessment)
```
:::

::: {.column width="60%"}

:::{.incremental}
* 3 variables, 12 observations, 36 values
* Each combination of name and assessment is a single measured observation
* Definition of a variable depends on the task
* E.g. what if we wanted to compare quiz and exam scores?
:::
:::

::::

## Keeping data tidy: tools 🔧

* Python: `pandas`, `polars`
* R: `tidyr`, `dplyr`, `polars` (not yet stable)

. . .

‼️ but cleaning and tidying your data is no substitute for good planning!

## Documentation

[TODO - move later?]

## Methodology and protocols

Methods for wet lab work get a lot of attention, but less so for computational work!

Key questions:

* What did you do?
* Why did you do it that way?

## Methodology and protocols

**What** did you do?

* Scripts - document the actual code you ran
* Computational environment - hardware, OS, package versions, random seeds
  * Python: `python -m venv`, `virtualenv`, `pyenv`,...
  * R: `renv`
  * Conda
  * Containers: Docker, Singularity,...
  
::: notes
compare to running ad hoc commands on commandline / ad hoc analyses
- e.g. renaming files, selecting columns/rows
- e.g. how did you make that plot?
:::

## Methodology and protocols

**Why** did you do it that way?

* Version control commit messages 
* Literate programming - combine code with text
  * Jupyter notebooks, RMarkdown, Quarto...
* Code reviews
* Architectural decision records

## Testing and controls

* Data validation
  * size and shape, data types, missing data
  * checksums
* Software testing
  * unit tests
  * snapshot testing - does this code still give the same output?
  * integration testing - does everything still work together?

## Automation

✨Automate the boring things✨

* Scripts for data cleaning / processing
* Combine scripts into workflows: run a series of data analysis steps in order
* Continuous integration - automatically run your tests

## Automation: tools 🔧

* Workflow managers
  * Nextflow: lots of public workflows for bioinformatics
  * Snakemake: Python-based, easy to create your own workflows
* Continuous integration
  * GitHub Actions: lots of templates

::: notes
Github Actions templates for e.g. building and running tests for a python package, for code linting and formatting, etc
:::

## Publishing

* FAIR principles
* persistent identifiers - how to get a DOI?
* licensing
* versioning




